Multidimensional information

### 2022-09-28 notebooks/Multidimensional_information.qmd
Update to the multidimensional information notebook

---
title: "Multidimensional information"
author: "Daniel Morillo"
format: html
bibliography: ../www/Multidimensional-parameters-MCLM.bib
editor: visual
---

@@

Hereby I am going to derive the multidimensional information, according to
@ackerman_creating_1994.

@@

@@

@@

## Components of $\mathbf{\beta}$

@@

@@

```{r}
rho     <- .5
sigma   <- c(1, rho, rho, 1) |> matrix(2)
sig_inv <- sigma |> solve()
```

@@

```{r}
a_item <- c(1, 0) |> matrix()
d_item <- 0
```

@@

```{r}
comp_vec <- c(0, 1) |> matrix()
```

@@

```{r}
library(magrittr)

ip_matrix <- sig_inv

@@

```{r}
A_matrix <- a_item |> t()

@@

As @ackerman_creating_1994 states (and I already knew) "for a single item or for
$n$ items that measure in exactly the same direction \[...\], Equation 7 (i.e.
the information matrix) is singular" (p. 259). This implies that I need at least
two items, and I want the information to be mostly aligned with one of the axes,
but they must not measure in the same direction. Therefore, I create a second
item, with near simple structure on the trait of the first axis.

```{r}
a_item_1 <- a_item
a_item_2 <- c(.95, .1) |> matrix()

@@

```{r}
error_var <- t(comp_cos_vec) %*% A_sq_inv %*% comp_cos_vec
comp_info <- 1 / error_var
```

The resulting multidimensional information coefficient for the direction
parallel to the second axis is `r comp_info`. Let's try to compute it in the
direction of the first axis:

```{r}
comp_vec     <- c(1, 0) |> matrix()
comp_vec_mod <- t(comp_vec) %*% ip_matrix %*% comp_vec |>
  drop() |>
  sqrt()
comp_cos_vec <- t(comp_vec) %*% ip_matrix %*% basis_matrix |>
  divide_by(comp_vec_mod * basis_vec_mods) |>
  matrix()

error_var <- t(comp_cos_vec) %*% A_sq_inv %*% comp_cos_vec
comp_info <- 1 / error_var
```

In the first axis the coefficient is `r comp_info`. This seems rather small when
one should expect a much higher information than in the second axis.

>>>>

---
title: "Multidimensional information"
author: "Daniel Morillo"
format:
  html:
    toc: true
    toc-location: left
    toc-depth: 4
bibliography: ../www/Multidimensional-parameters-MCLM.bib
editor: visual
---

```{r setup}
#| include: false
options(digits = 5)
```

@@

Hereby I am going to try to derive the multidimensional information, according
to @ackerman_creating_1994.

@@

## Multidimensional information

@@

## Linear composite $\mathbf{\beta}$

@@

### Components of $\mathbf{\beta}$

@@

## Example with a correlated bidimensional space

@@

```{r def-corr}
rho <- .5
```

```{r def-covariance}
sigma   <- c(1, rho, rho, 1) |> matrix(2)
sig_inv <- sigma |> solve()
```

@@

```{r def-item}
a_item <- c(1, 0) |> matrix()
d_item <- 0
```

@@

```{r def-composite}
comp_vec <- c(0, 1) |> matrix()
```

### Computation in the latent space {#latent-space}

@@

```{r def-inner-product}
ip_matrix <- sig_inv
```

```{r compute-composite-components}
library(magrittr)

@@

```{r compute-discr-matrix-attemt}
A_matrix <- a_item |> t()

@@

As @ackerman_creating_1994 states "for a single item or for $n$ items that
measure in exactly the same direction \[...\], Equation 7 (i.e. the information
matrix) is singular" (p. 259). This I already knew, but hadn't realized it was
the case; therefore, my previous assertion that "in the direction of the other
trait axis, the information should be higher the more positive the correlation
is" is actually false (for one item). Therefore, I need at least two items, and
I want the information to be mostly aligned with one of the axes, but they must
not measure in the same direction. Thus I create a second item, with near simple
structure on the trait of the first axis.

```{r def-items}
a_item_1 <- a_item
a_item_2 <- c(1.5, .1) |> matrix()

@@

```{r compute-info-axis2}
error_var  <- t(comp_cos_vec) %*% A_sq_inv %*% comp_cos_vec
comp_info2 <- 1 / error_var
```

The resulting multidimensional information coefficient for the direction
parallel to the second axis is `r comp_info2`. Let's try to compute it in the
direction of the first axis:

```{r compute-info-axis1}
comp_vec     <- c(1, 0) |> matrix()
comp_vec_mod <- t(comp_vec) %*% ip_matrix %*% comp_vec |>
  drop() |>
  sqrt()
comp_cos_vec <- t(comp_vec) %*% ip_matrix %*% basis_matrix |>
  divide_by(comp_vec_mod * basis_vec_mods) |>
  matrix()

error_var  <- t(comp_cos_vec) %*% A_sq_inv %*% comp_cos_vec
comp_info1 <- 1 / error_var

ratio <- comp_info1 / comp_info2
```

In the first axis the coefficient is `r comp_info1`. This seems rather small
when one should expect a much higher information than in the second axis; the
ratio in the two directions is `r ratio`.

#### Test with higher correlation

Common sense tells us that the more aligned the axes are, the more similar the
information will be on the two dimensions parallel with the two axes. To test
this commonsense assumption, we try the previous computations again with a
higher correlation.

```{r test-higher-corr}
# Correlation and covariance matrix
rho       <- .93
sigma     <- c(1, rho, rho, 1) |> matrix(2)
sig_inv   <- sigma |> solve()
ip_matrix <- sig_inv

# Basis vectors
basis_vec_mods <- t(basis_matrix) %*% ip_matrix %*% basis_matrix |>
  diag() |>
  sqrt()

# Composite cosine vectors
comp_vecs     <- diag(2)
comp_vecs_mod <- t(comp_vecs) %*% ip_matrix %*% comp_vecs |>
  diag() |>
  sqrt()
comp_cos_vecs <- t(comp_vecs) %*% ip_matrix %*% basis_matrix |>
  divide_by(comp_vecs_mod * basis_vec_mods)

# Error variance and information in the composite directions
error_var <- t(comp_cos_vecs) %*% A_sq_inv %*% comp_cos_vecs |> diag()
comp_info <- 1 / error_var

ratio <- comp_info[1] / comp_info[2]
```

The information in the two axes is now `r comp_info[1]` and `r comp_info[2]`,
being their ratio `r ratio`. This fulfills the assumption we wanted to test.
However, it doesn't make much sense that the information in the direction of the
first axis has decreased.

### Computation in the test space

We are going to test the same assumption, but this time in the test space. This
implies using the covariance matrix as inner product matrix, instead of its
inverse.

```{r}
rho <- .5
```

We try first with a correlation of `r rho`.

```{r compute-test-space}
# Covariance matrix
sigma     <- c(1, rho, rho, 1) |> matrix(2)
ip_matrix <- sigma

# Basis vectors
basis_vec_mods <- t(basis_matrix) %*% ip_matrix %*% basis_matrix |>
  diag() |>
  sqrt()

# Composite cosine vectors
comp_vecs     <- diag(2)
comp_vecs_mod <- t(comp_vecs) %*% ip_matrix %*% comp_vecs |>
  diag() |>
  sqrt()
comp_cos_vecs <- t(comp_vecs) %*% ip_matrix %*% basis_matrix |>
  divide_by(comp_vecs_mod * basis_vec_mods)

# Error variance and information in the composite directions
error_var <- t(comp_cos_vecs) %*% A_sq_inv %*% comp_cos_vecs |> diag()
comp_info <- 1 / error_var

ratio <- comp_info[1] / comp_info[2]
```

The information in both axes directions would be `r comp_info[1]` and
`r comp_info[2]`, and their ratio `r ratio`.

```{r}
rho <- .93
```

Now we compute the same result, but with a correlation of `r rho` as we did
before.

```{r compute-test-space-high-corr}
# Covariance matrix
sigma     <- c(1, rho, rho, 1) |> matrix(2)
ip_matrix <- sigma

# Basis vectors
basis_vec_mods <- t(basis_matrix) %*% ip_matrix %*% basis_matrix |>
  diag() |>
  sqrt()

# Composite cosine vectors
comp_vecs     <- diag(2)
comp_vecs_mod <- t(comp_vecs) %*% ip_matrix %*% comp_vecs |>
  diag() |>
  sqrt()
comp_cos_vecs <- t(comp_vecs) %*% ip_matrix %*% basis_matrix |>
  divide_by(comp_vecs_mod * basis_vec_mods)

# Error variance and information in the composite directions
error_var <- t(comp_cos_vecs) %*% A_sq_inv %*% comp_cos_vecs |> diag()
comp_info <- 1 / error_var

ratio <- comp_info[1] / comp_info[2]
```

The information would be now `r comp_info[1]` and `r comp_info[2]` for the first
and second axis direction, respectively, and their ratio `r ratio`.

### Interpretation of the differences in the results in both spaces

The information in the direction of the first axis decreases also, as happened
when computing the directions in the latent space. However, the information in
the direction of the second axis increases, while when computing it in the
latent space it also decreased. It makes more sense to assume that the direction
would increase in the second axis direction with an increase in the correlation,
as it gets more aligned with the first axis.

On the other hand, exploring the cosines that determine the composite direction
one can see that, when using the latent space, they get more aligned **in the
opposite direction** when the correlation increases. This makes me think that,
in this case, the information is more evenly distributed across all direction in
the latent space. However, what I would expect is that the information get more
"concentrated" in a single direction.

### Correction about the computation of the information matrix

There is also one last thing when computing the information with two (or more)
items: I have assumed that the $PQ^{-1}$ factor can be factored out of the
matrix (see section [Computation in the latent space](#latent-space) above).
However, this is wrong. Its value is different for each item, and thus
it must be taken into account when computing the sum of the information for all
the items. This implies that the information matrix cannot be computed as
$\mathbf{A}^T\mathbf{A}PQ$ as I assumed before, for more than one item.
